---
title: "Estimate cellular composition from blood DNA methylation data"
author: "Eilis Hannon"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\EstimateBloodCellularComposition{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This vignette demonstrates how to use the reference DNA methyation data to estimate cellular proportions from whole blood DNA methylation data. The reference data were generated with the Illumina Infinium MethylationEPIC BeadChip and includes CD4 T cells, CD8 T cells, B cells, Monocytes and Granulocytes. Th output is a matrix with one column per cell type and one row per sample where the values represent the estimated cellular proportion for that sample and that cell type. 


### Install the package

The commands below will install the package directly from github. NB it assumes minfi is already installed. 

```{r,eval=FALSE}

install.packages("devtools")
library("devtools")
install_github("ejh243/EPICBloodCompCalc")
library(EPICBloodCompCalc)
library(minfi)

```

### Estimate cellular composition

To etsimate cellular composition you need to load your data in from idats files to an RGChannel Set object. This code will do this.


```{r,eval=FALSE}
samplesToLoad<- ## list of samples typically SentrixID_SentrixPosition for which you have two idat files
setwd() ## change to directory where idat files are located
RGSet <- read.metharray(samplesToLoad) ## load from idat files

```

To estimate cellular composition for these samples run:


```{r,eval=FALSE}
samplesToLoad<- ## list of samples typically SentrixID_SentrixPosition for which you have two idat files
setwd() ## change to directory where idat files are located
RGSet <- read.metharray(samplesToLoad) ## load from idat files

```



```{r,eval=FALSE}
## to estimate cellular proportions for data generated with the EPIC array
counts<-estimateCellCountsEPIC(RGSet, EPIC=TRUE, cellTypes=c("Bcells", "CD4Tcells", "CD8Tcells", "Monocytes", "Granulocytes"))

```

You can adjust the 'cellTypes' argument to select a subset of cell types to estimate (for example if you have PBMCs and don't want to estimate the proportion of granulocytes).


By default the package assumes that you are going to supply DNA methylation data profiled with the EPIC array, if you infact have data from the 450K array change the 'EPIC' argument to FALSE as shown here

```{r,eval=FALSE}
## to estimate cellular proportions for data generated with the 450K array
counts<-estimateCellCountsEPIC(RGSet, EPIC=FALSE, cellTypes=c("Bcells", "CD4Tcells", "CD8Tcells", "Monocytes", "Granulocytes"))

```
